Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques

### 2019-04-15 paper/paperV1.tex
Proofread 2

In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm used until recently, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Specifically, the axis of pleasantness is found as the most explanatory of subjective assessments, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}. Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments. While the taxonomy used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\

Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps. With the advent of the Internet of Things and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017}, it may be feasible to predict pleasantness or other perceptual factors from acoustic measurements. Bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is an interesting and ongoing area of research. Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level. Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments. Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes. These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited.\\

In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness. In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. However, large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature. In order to circumvent this issue, we propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017}.\\

The contributions of this paper are thus three-fold:
\begin{enumerate}
  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception
  \item Propose numerical means for predicting the time of presence of sound sources from raw acoustic data
  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
\end{enumerate}

First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use in the context of pleasantness prediction. Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Estimations of pleasantness using these predictions are next compared to existing models from physical indicators in Section~\ref{sec:results}.

@@

A corpus is first created to validate the use of simulated sound scenes in the development of pleasantness prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception. The resulting 45~s extracts are replicated to study perceptual differences between recorded and simulated scenes. The recordings are first annotated in terms of the following properties:

@@

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Because of the lack of large isolated sources datasets, we choose to restrict the scene composition to 3 sources: \textit{traffic}, \textit{human voices} and \textit{birds}. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in the form of normal distributions,
\item The inter-onset of event occurrences in seconds, also in the form of normal distributions.
\end{itemize}

@@

Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}

Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. A random pre-generated listening order is used to attenuate ordering effects, although all participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15) to calibrate their answers. For the remainder of the corpus all participants experience different scene ordering. Participants can only listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\

@@

A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.

@@

Participants evaluate each scene on criteria coded on 11-point discrete scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus, and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace} compares the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace} using arrows. The standard deviation of assessments is represented by ellipses for one scene. The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, not shown here for clarity. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

% FIGURE
\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.55\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_p1.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/pca_sim.eps}
    \end{subfigure}
    \caption{Principal components analysis (first two components) of the four general questions at the assessment level (n=252). Each dot represent the projection of one scene on the resulting space. The components space is distorted compared to the literature.}\label{fig:pspace}
\end{figure}

@@

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. To provide a baseline for pleasantness prediction the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\

The simulation process outputs ground truth source contributions in the form of separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

@@

To train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
\begin{itemize}
\item The development set containing 400 45~s scenes (total 5 hours) which is used during the training process,
\item The evaluation set containing 200 45~s scenes (total 2.5 hours) which is used to compute the performance of the trained model and its generalization capabilities.
\end{itemize}

The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}. Only traffic, human voices and birds sources compose the sound scenes due to the low number of available samples and recording annotations for other sources. The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range. Inputs are then obtained by splitting the resulting spectrograms into frames with a 1~s duration, which is considered a relevant scale for perception. The input frames of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values obtained before averaging along time can be used as weak labels. Alternatively, it is possible to perform sound level estimation using ground truth labels. Though, this regression task is difficult to learn compared to source detection, and will not be discussed here as it did not result in better time of presence estimations in our experiments.\\

@@

We first study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in the previous section and for the relations to source contributions. Pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

@@

Table~\ref{tab:physm} compares the performance of the three models. Expectedly, $\hat P_{\varphi 1}$ outperforms both baselines. Though, compared to the best perceptual models its explained variance is about 9\% lower. Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 10-point scale. In $\hat P_{\varphi 2}$ introducing the $L_{10}-L_{90}$ emergence has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\

@@

The performance of the source detection deep learning architecture presented in Section~\ref{sec:methods_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}. The overall presence detection accuracy is 92.11\%. The model performs similarly well for the three sources, ranging from 90.08\% for birds to 94.09\% for traffic. Regarding the type of errors, they are equally split between false positives and false negatives. Though, these rates vary depending on the source. Traffic has a very low false positive rate at 2.46\% and high false negative rate, while birds display the highest false positive rate at 12.30\%. This is expected given the spectral content related to these sources. Traffic is usually located in lower frequencies than voice or bird events, but may contain high frequency components mistaken for birds by the model. The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.\\

@@

Next, the capacity of these predictions to give relevant pleasantness estimates is studied. For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:results_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. First, pleasantness estimations are equally performant using the model's predictions or the ground truth labels , with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

@@

The physical model on this corpus includes only the contribution of bird sources. This result is attributed to the overall low diversity of the studied corpus:
\begin{itemize}
\item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process. Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic. It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
\item The diversity of available isolated source extracts for scene simulation is rather low. Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness. These properties result in no significant contribution of voice events to pleasantness for this corpus.
\end{itemize}
Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness. The trained deep learning architecture already displays few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\

Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels. Thus, in some cases the deep learning model's predictions are considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\

The detection-based pleasantness predictions are of lower quality on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy. This is expected as the architecture is not trained on complex scenes. Though, interestingly pleasantness estimates are still better on average than on the corpus of simulated scenes. The detection model is thus quite robust to scenes with additional sources as perturbators. A full study with a larger evaluation corpus would be needed to further validate the model's robustness to diverse real-life scenarios.

@@

Subsequently, a state-of-the-art deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction in urban environments, leads to better estimates than previous models from acoustic indicators.\\

Future work will focus on the robustness of the source detection model, which heavily depends on the quantity and quality of data available for scene simulation. Furthermore, this methodology can be adapted to specific environments by extending or refining the source taxonomy in both the scene simulation process and detection model. For example, additional transport-related sources such as train and plane activity could be introduced, and a difference could be made depending on voice expressivity or bird species.

>>>>

In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Specifically, the axis of pleasantness emerges as the first dimension in subjective assessments of soundscapes, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}. Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments. While the exact taxonomy of sources used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\

Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps. Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost. Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.\\

Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level. Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments. Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes. These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited. In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.\\

In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature. We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
\end{enumerate}

First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use for the prediction of pleasantness. Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Estimations of pleasantness using these predictions are next compared to existing models from physical indicators in Section~\ref{sec:results}.

@@

A corpus is first created to validate the use of simulated sound scenes in the development of pleasantness prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings, 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception. The resulting 45 second extracts are replicated to study perceptual differences between recorded and simulated scenes. The recordings are first annotated in terms of the following properties:

@@

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}. We choose to restrict the scene composition to these 3 sources. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in terms of normal distributions,
\item The inter-onset of event occurrences in seconds, also in terms of normal distributions.
\end{itemize}

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}

Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15) to calibrate their answers. A random listening order is generated for each subject to attenuate ordering effects for the remaining of the test. Participants can only listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\

@@

A total of 23 students aged from 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.

@@

Participants evaluate each scene on criteria using 11-point Likert scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the judgements of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus. Starting with the assessments of each participant as separate clusters, new clusters are formed iteratively where the average euclidean distance between all pairs within the two merged clusters at each step is minimal. and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

@@

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace_rec} as dots. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented by ellipses for one scene. The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/pca_p1.eps}
    \caption{Principal components analysis of the four general questions for scenes with real scenarios at the assessment level (n=575). Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown by ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/pca_sim.eps}
    \caption{Principal components analysis of assessments for the simulated scenes (red dots) at the assessment level (n=1725), with projections of recorded and replicated scenes (blue dots).}\label{fig:pspace_sim}
\end{figure}

@@

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\

The simulation process outputs ground truth source contributions as separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

@@

To train deep learning architectures for source recognition, a larger corpus is constructed. It is composed of two datasets of simulated scenes:
\begin{itemize}
\item The development set of 5 hours containing 400 scenes of 45~s each, which is used during the training process,
\item The evaluation set of 2.5 hours containing 200 scenes of 45~s each, which is used to compute the performance of the trained model and its generalization capabilities.
\end{itemize}

The generation procedure is the same for both datasets, and is equivalent to that of the simulated sub-corpus considered for the perceptual test in Section~\ref{sec:data_corp}. Only traffic, human voices and birds sources compose the sound scenes. The isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora both the scenarios and samples are different.\\

The developped model should extract relevant source information from a representation of the audio signal. Typically, spectral representations such as mel bands are preferred to the raw audio waveform because of the regularities they underline in the signal. Here, the third-octave spectrum is considered as it is commonly used in acoustic monitoring applications~\cite{ardouin2018, gontier2017}. Third-octave spectra are computed for 125~ms frames and 29 frequency bands in the $20~Hz - 12.5~kHz$ range. Inputs are then obtained by splitting the resulting spectrograms into texture windows of 1~s duration, which is considered a relevant scale for perception. The spectral blocks of dimension 29x8 are then processed independently in this study.\\

Target outputs are associated to each input frame to train the model. In~\cite{gontier2018} the time of presence of traffic, voices and birds was successfully approximated by the $\hat T_s(\alpha, \beta)$ indicator. Thus, the source presence values given by this indicator can be used as weak perceptual time of presence labels. Alternatively, it is possible to perform sound level estimation using ground truth labels provided by the simulation process before computing $\hat T_s(\alpha, \beta)$. Though, training a model for sound level regression is more difficult than for source detection as it requires the architecture to perform implicit source separation. This will not be discussed here as it did not result in better time of presence estimations in our experiments.\\

@@

We first study the relations between perceptual scales with respect to existing work. Table~\ref{tab:percc} shows the Pearson's correlation coefficients at the scene level, that is on the mean responses for each scene. The resulting values are consistent with the literature~\cite{aumond2017, gontier2018}, both between general scales as discussed in Section~\ref{sec:data_val} and for the relations to source contributions. Indeed, pleasantness (P) is mainly influenced negatively by overall loudness and traffic and positively by birds. In previous studies a small positive contribution of voices to pleasantness was found, while no direct relation is visible from the data gathered. This can be explained by the choice of speech samples used during generation, which consist of read audiobooks extracts and thus may sound unnatural in the considered urban environments. Liveliness (L) is here negatively though weakly (r<0.4) correlated with the presence of birds, while no significant relation was found in previous studies.\\

@@

Table~\ref{tab:physm} compares the performance of the three models. Expectedly, $\hat P_{\varphi 1}$ outperforms both baselines. Though, compared to the best perceptual models its explained variance is about 9\% lower. Its RMSE is also significantly higher, but acceptable considering the average standard deviation of pleasantness assessments of 1.77 on a 10-point scale. Introducing the $L_{10}-L_{90}$ emergence in $\hat P_{\varphi 2}$ has no significant impact on predictions, using only the $L_{50}$ results in the same performance.\\

@@

The performance of the source detection deep learning architecture presented in Section~\ref{sec:methods_deep} on the evaluation dataset is summarized in Table~\ref{tab:perf_cmp}. The overall presence detection accuracy is 92.11\%. The model performs similarly well for the three sources, ranging from 90.08\% for birds to 94.09\% for traffic. Regarding the type of errors, they are equally split between false positives and false negatives. Though, these rates vary depending on the type of source. Traffic has a very low false positive rate at 2.46\% and high false negative rate, while birds display the highest false positive rate at 12.30\%. This is expected given the spectral content related to these sources. Traffic is usually located in lower frequencies than voice or bird events, but may contain high frequency components mistaken for birds by the model. The resulting error on the time of presence is 0.12 RMSE overall on a 0-1 scale, and is close for the three sources.\\

@@

Next, the capacity of these predictions to give relevant pleasantness estimates is studied. For this analysis, only the perceptual experiment is used as no assessments are available on the evaluation dataset. To evaluate the model's robustness to the increased polyphony and source complexity of scenes in real-life scenarios, the corpus is further split in two parts:
\begin{enumerate}
\item The 25 recorded and replicated scenes, which contain additional sources not present in the development or evaluation datasets simulated for the optimization of deep learning models,
\item The 75 simulated scenes that are obtained from the same simulation process as both the development and evaluation dataset.
\end{enumerate}
Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{\varphi 1}$ model presented in Section~\ref{sec:results_phys}. Thus, only the outputs corresponding to the presence of birds are used. The results are compared to estimates using ground truth $\hat T_s(\alpha, \beta)$ labels in Table~\ref{tab:pppred}. First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with an RMSE of about 0.84 and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates. Labels from the detection model result in higher errors on the first sub-corpus containing recorded and replicated scenes, with sources not seen during training. In both cases the error is still lower on the first sub-corpus, indicating that the detection model generalizes well to new content and sources.\\

@@

The physical model on this corpus includes only the contribution of bird sources. This result is attributed to the overall low diversity of the studied corpus:
\begin{itemize}
\item Only traffic, voice and bird sources are present in simulated sound scenes, as a result global sound level measurements tend to be correlated with the presence of traffic in the simulation process. Quieter environments such as parks and quiet streets are less likely to have continuous traffic, while high sound levels in busy streets are always due to busy traffic. It is expected that including other sources such as construction noises and other transport-related contributions in such environments would lower the observed correlation.
\item The diversity of available isolated source extracts for scene simulation is rather low. Particularly, voice events are recordings of read english with low variations in speaker and consistently neutral expressiveness. These properties result in no significant contribution of voice events to pleasantness for this corpus.
\end{itemize}
Thus, in real life scenarios where detection models are applied to sensor networks, indicators related to traffic and voice sources will become relevant for the prediction of pleasantness. The trained deep learning architecture already displays only few detection errors for these sources, and can be extended to any additional source provided enough data is available.\\

Pleasantness estimates using the proposed source detection architecture are similar to those using ground truth $\hat T_s(\alpha, \beta)$ labels on the simulated corpus. This indicates that detection errors found in Table~\ref{tab:perf_cmp} do not impact pleasantness prediction on average. Since the labels are extracted from a masking model approximating the perceptual time of presence, they can be considered as "weak" labels. Thus, the deep learning model's predictions are in some cases considered erroneous but correlate better to perceptual assessments than the corresponding ground truth labels of $\hat T_s(\alpha, \beta)$.\\

The detection-based pleasantness predictions are of lower quality on the corpus containing recorded and replicated scenes, which feature higher polyphonies and a richer source taxonomy. This is expected as the architecture is not trained on complex scenes. Though, pleasantness estimates are interestingly still better on average than on the corpus of simulated scenes. The detection model is thus quite robust to scenes with additional sources as perturbators.

@@

Subsequently, a state-of-the-art deep learning architecture for source presence detection is developed that, when applied to pleasantness prediction from acoustic indicators in urban environments, leads to better estimates than previous state-of-the-art models.\\

Future work will focus on the robustness of the source detection model, which heavily depends on the quantity and quality of data available for scene simulation. Together with additional recordings, data augmentation techniques for machine learning will be studied to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching. Furthermore, this methodology can be adapted to specific environments by extending or refining the source taxonomy in both the scene simulation process and detection model. For example, additional transport-related sources such as train and plane activity could be introduced, and a difference could be made depending on voice expressivity or bird species.
